---
title: "Data Wrangling Assignment 1"
author: "Alexander Uitendaal, Milad Ehsan(10732918) , Daan Uittenhout() and Isabelle Slot(10676848)"
date: "27/09/2018"
output: html_document
---
```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## Loding the data

First instal the mlbench package and add it to the Working directory llibary
```{r}
install.packages("mlbench") 
rm(list=ls())         # remove working memory
library("mlbench")    #install the right package
data("BreastCancer")  #install the right dataset: BreastCancer
View(BreastCancer)
```
## Preprocessing the data

Check the data on missing values and omit these values
```{r}
## Missing values 
# Check whether there are missing values
missing <- is.na(BreastCancer) 
summary(missing) 
# There are missing values, so delete these:
BreastCancer2<-na.omit(BreastCancer)
summary(BreastCancer2)

```
Conclusion after checking the data, only 16 out of 700 (+/- 2%) observations where lost, so imputing is not necessary because 
only very little information is lost because of the missing values 

## Visualizing the data stage 1


```{r}
## Visualise the variables and see their relation with the Class type of the Cancer
## Also look for outliers, 

# instal the nececary packages and add them to the libary
install.packages("ggplot2")
library("ggplot2")

"Variable 1 : cell thickness"
# Plot the a frequency relation between cell thikness and breastcancer 
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Cl.thickness,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Clump Thickness")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

"Variable 2 : cell size"
# Plot the a frequency relation between cell size and breastcancer 
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Cell.size,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Cell Size")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

"Variable 3 : cell shape"
# Plot the a frequency relation between cell shape and breastcancer 
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Cell.shape,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Cell Shape")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

"Variable 4 : cell marginal adhesion"
# Plot the a frequency relation between cell marginal adhesion and breastcancer 
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Marg.adhesion,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Marginal Adhesion")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

"Variable 5 : Single Epithelial Cell Sizehesion"
# Plot the a frequency relation between Single Epithelial Cell Sizehesion and breastcancer 
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Epith.c.size,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Single Epithelial Cell Size")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

"Variable 6 : Bare Nuclei"
# Plot the a frequency relation between Bare Nuclei and breastcancer 
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Bare.nuclei,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Bare Nuclei")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

" Variable 7 bland Chromatin"

ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Bl.cromatin,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Bland Chromatin")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

"Variable 8 Normal Nucleoli"

ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Normal.nucleoli,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Normal Nucleoli")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 

" Variable 9 Mitoses"

ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Mitoses,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Mitoses")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
# No outliers spotted 


```
All the variables are on an ordinal scale from 1 to 10. Therefore there are no clear   outliers in the data to be spotted 
For all the variable seems  that the higher the value, the more likely the obervervations 
class type is 'malignant' 
```{r}
"make a correlation matrix colored, so we can have a quick overview of strong and weak correlating variables. First make the variables numeric so we can plot the correlation matrix "

#install.packages("corrplot")
#install.packages("taRifx")

library( "taRifx")
library("corrplot")

breastcancer3 <- BreastCancer2[2:10]
breastcancer3 <- sapply( breastcancer3, as.numeric )
correlation_matrix <- cor(breastcancer3)
correlation_plot <- corrplot(correlation_matrix, method = 'color',type = 'upper', addCoef.col = "black")
# strong correlation between cell shape and cell size 
# all variables have a positive correlation
```

```{r}
#contingency table Clump thickness
install.packages("descr")
library("descr")
tabel1 <-CrossTable(BreastCancer2$Class, BreastCancer2$Cl.thickness,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel1)
#shows that the thinkess of the Clump is not bigger than 8 for a benign tumor. 
#Moreover, if the thickness of the clump is smaller than 6, the tumor is mostly benign, vicaversa for thickness >= 6 


#contingency table  Cell Size
tabel2 <-CrossTable(BreastCancer2$Class, BreastCancer2$Cell.size,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel2)
#shows that, except for a few unique cases, the cell size usually does not get bigger than 4 for a benign tumor. 

#contingency table Cell Shape
tabel3 <-CrossTable(BreastCancer2$Class, BreastCancer2$Cell.shape,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel3)
#cell shape is decreasing for bening tumors. Malignant cases show a less clear pattern.

#contingency table Marginal Adhesion
tabel4 <-CrossTable(BreastCancer2$Class, BreastCancer2$Marg.adhesion,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel4)
# except for one benign tumor, the marginal adhesion has a decreasing pattern for benign tumors and doesnt get bigger than 6.
# the malignant cases show a less abrupt decrease and do not really show a consistent pattern. 

#contingency table  Epith Size
tabel5 <-CrossTable(BreastCancer2$Class, BreastCancer2$Epith.c.size,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel5)
#  benign cases are decreasing, whereas the malignant cases do not really show a consistent pattern.

#contingency table Bare Nuclei
tabel6 <-CrossTable(BreastCancer2$Class, BreastCancer2$Bare.nuclei,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel6)
# benign cases are decreasing, malignant cases are quite consistent, except for the great amountin class 10. 
#One can say that there is a high  probability of a malignant tumor when the Bare nuclei class is 10 and visa versa for class 1.

#contingency table Bl Cromatin
tabel7 <-CrossTable(BreastCancer2$Class, BreastCancer2$Bl.cromatin,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel7)
#decreasing pattern for benign cases, with no Bland Chromatin bigger than 7 for a benign case. 


#contingency table Normal Nucleoli
tabel8 <-CrossTable(BreastCancer2$Class, BreastCancer2$Normal.nucleoli,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel8)
#benign class shows a decreasing pattern, with no Normal nucleoli bigger than 8
#

#contingency table Mitoses
tabel9 <-CrossTable(BreastCancer2$Class, BreastCancer2$Mitoses,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel9)
#decreasing pattern for benign cases. Allmost all benign cases have a Mitoses class of 1. 
#The malignant cases are also decreasing, but mitoses bigger than 1 are less rare.
```

