---
title: "Data Wrangling Assignment 1"
author: "Alexander Uitendaal, Milad Ehsan(10732918) , Daan Uittenhout(11057777) and Isabelle Slot(10676848)"
date: "27/09/2018"
output: html_document
---
```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## Loading the data

First instal the mlbench package and add it to the working directory library
```{r eval = FALSE}
install.packages("mlbench") 
rm(list=ls())         # remove working memory
library("mlbench")    #install the right package
data("BreastCancer")  #install the right dataset: BreastCancer
View(BreastCancer)
```
## Preprocessing the data

Check the data on missing values and omit these values
```{r}
## Missing values 
# Check whether there are missing values
missing <- is.na(BreastCancer) 
summary(missing) 
# There are missing values, so delete these:
BreastCancer2<-na.omit(BreastCancer)
summary(BreastCancer2)

```
Conclusion after checking the data, only 16 out of 700 (+/- 2%) observations where lost, so imputing is not necessary because 
only very little information is lost because of the missing values 

## Visualizing the data

### Visualizing the variables and see their relation with the Class type of the Cancer
### *Also look for outliers*

```{r}
# instal the nececary packages and add them to the libary
install.packages("ggplot2")
library("ggplot2")
```

#### Variable 1 : cell thickness
#### Plot the a frequency relation between cell thikness and breastcancer 

```{r}
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Cl.thickness,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Clump Thickness")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
```
#### *No outliers spotted*

#### Variable 2 : cell size
#### Plot the a frequency relation between cell size and breastcancer 
```{r}
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Cell.size,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Cell Size")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))

```
#### *No outliers spotted*


#### Variable 3 : cell shape
#### Plot the a frequency relation between cell shape and breastcancer 


```{r}

ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Cell.shape,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Cell Shape")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
 
```
#### *No outliers spotted*

#### Variable 4 : cell marginal adhesion
#### Plot the a frequency relation between cell marginal adhesion and breastcancer 


```{r}

ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Marg.adhesion,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Marginal Adhesion")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
```
#### *No outliers spotted*

#### Variable 5 : Single Epithelial Cell Sizehesion
#### Plot the a frequency relation between Single Epithelial Cell Sizehesion and breastcancer 
```{r}
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Epith.c.size,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Single Epithelial Cell Size")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
```
#### *No outliers spotted*

#### Variable 6 : Bare Nuclei
#### Plot the a frequency relation between Bare Nuclei and breastcancer 
```{r}
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Bare.nuclei,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Bare Nuclei")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))

```
#### *No outliers spotted*

#### Variable 7 bland Chromatin"
#### Plot the a frequency relation between bland Chromatin and breastcancer 

```{r}
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Bl.cromatin,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Bland Chromatin")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))
```
#### *No outliers spotted*

#### Variable 8 Normal Nucleoli
#### Plot the a frequency relation between Normal Nucleoli and breastcancer 
```{r}
ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Normal.nucleoli,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Normal Nucleoli")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))

```
#### *No outliers spotted*

#### Variable 9 Mitoses
#### Plot the a frequency relation between Mitoses and breastcancer 
```{r}
" Variable 9 Mitoses"

ggplot() +
  geom_bar(data = BreastCancer2,
           aes(x = Mitoses,
               fill = BreastCancer2$Class),
           position = "Stack") + scale_x_discrete("Mitoses")+scale_y_continuous("Frequency")+
  guides(fill=guide_legend(title="Class type"))+scale_fill_manual(values=c("red", "blue"))

```
#### *No outliers spotted*

### conclusion
* All the variables are on an ordinal scale from 1 to 10. Therefore there are no clear outliers in the data to be spotted 
* For all the variable seems that the higher the value, the more likely the obervervations class type is 'malignant'

## Correlation and Contingency


### Correlation Matrix
*The correlation matrix was colored, so there is a quick overview of strong and weak correlating variables. The variables are made numeric so these could be plotted in the correlation matrix* 


```{r}
#instal the right packages and them to the working directory library
install.packages("corrplot")
install.packages("taRifx")
library( "taRifx")
library("corrplot")

breastcancer3 <- BreastCancer2[2:10]
breastcancer3 <- sapply( breastcancer3, as.numeric )
correlation_matrix <- cor(breastcancer3)
correlation_plot <- corrplot(correlation_matrix, method = 'color',type = 'upper', addCoef.col = "black")
```
### Conclusion
* Strong correlation between cell shape and cell size 
* All variables have a positive correlation 

## Contingency Tables

### contingency table Clump thicknes
```{r}
install.packages("descr")
library("descr")
tabel1 <-CrossTable(BreastCancer2$Class, BreastCancer2$Cl.thickness,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel1)
```
#### Analysis
* The table shows that the thinkess of the Clump is not bigger than 8 for a benign tumor. 
* The thickness of the clump is smaller than 6, the tumor is mostly benign, vicaversa for thickness (>= 6).

### Contingency table Cell Size
```{r}

tabel2 <-CrossTable(BreastCancer2$Class, BreastCancer2$Cell.size,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel2)
```

#### Analysis
* The table shows that, except for a few unique cases, the cell size usually does not get bigger than 4 for a benign tumor. 

### Contingency table Cell Shape

```{r}
tabel3 <-CrossTable(BreastCancer2$Class, BreastCancer2$Cell.shape,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel3)
```
#### Analysis
* Cell shape is decreasing for bening tumors. Malignant cases show a less clear pattern.

### Contingency table Marginal Adhesion

```{r}
tabel4 <-CrossTable(BreastCancer2$Class, BreastCancer2$Marg.adhesion,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel4)
```
#### Analysis
* Except for one benign tumor, the marginal adhesion has a decreasing pattern for benign tumors and doesnt get bigger than 6.
* The malignant cases show a less abrupt decrease and do not really show a consistent pattern. 

### Contingency table  Epith Size
```{r}
tabel5 <-CrossTable(BreastCancer2$Class, BreastCancer2$Epith.c.size,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel5)
```
#### Analysis
* Benign cases are decreasing, whereas the malignant cases do not really show a consistent pattern.

### Contingency table Bare Nuclei
```{r}
tabel6 <-CrossTable(BreastCancer2$Class, BreastCancer2$Bare.nuclei,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel6)
```
#### Analysis
* Benign cases are decreasing, malignant cases are quite consistent, except for the great amountin class 10. 
* One can say that there is a high  probability of a malignant tumor when the Bare nuclei class is 10 and visa versa for class 1.

### Contingency table Bl Cromatin
```{r}

tabel7 <-CrossTable(BreastCancer2$Class, BreastCancer2$Bl.cromatin,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel7)
```
#### Analysis
* Decreasing pattern for benign cases, with no Bland Chromatin bigger than 7 for a benign case. 


### Contingency table Normal Nucleoli
```{r}
tabel8 <-CrossTable(BreastCancer2$Class, BreastCancer2$Normal.nucleoli,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel8)
```
#### Analysis
* Benign class shows a decreasing pattern, with no Normal nucleoli bigger than 8

### Contingency table Mitoses
```{r}
tabel9 <-CrossTable(BreastCancer2$Class, BreastCancer2$Mitoses,prop.c = FALSE,prop.chisq = FALSE, prop.t = FALSE)
print(tabel9)
```
#### Analysis
* Decreasing pattern for benign cases. Allmost all benign cases have a Mitoses class of 1. 
* The malignant cases are also decreasing, but mitoses bigger than 1 are less rare.


